---
title: 'Assignment #2'
author: "Sameer Shankar"
date: "2/14/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

(a) 

$X_t = \frac{2}{5}X_{t-1} + \frac{1}{4}X_{t-2} - \frac{1}{10}X_{t-3} + Z_t$

$X_t - \frac{2}{5}X_{t-1} - \frac{1}{4}X_{t-2} + \frac{1}{10}X_{t-3} = Z_t$

$X_t - \frac{2}{5}BX_t - \frac{1}{4}B^2X_t + \frac{1}{10}B^3X_t = Z_t$

So,

$(1 - \frac{2}{5}B - \frac{1}{4}B^2 + \frac{1}{10}B^3)X_t = Z_t$

Therefore, we solve the following for B,

$1 - \frac{2}{5}B - \frac{1}{4}B^2 + \frac{1}{10}B^3 = 0$

Using the polyroot() function, we can find the roots for the equation above,

```{r characteristic polynomial, include=TRUE}
polyroot(z = c(1, -0.4, -0.25, 0.1))
```

We know that an AR(p) process $\phi(B)X_t = Z_t$ is stationary if and only if
the roots of the characteristic polynomial $\phi(B)$ lie outside the unit circle
in the complex plane.

Our roots are $B=2-0i, B=-2+0i, B=2.5+0i$. All of these roots clearly lie
outside the unit circle, which means {$X_t$} is stationary.

(b)

The Yule-Walker equations are,

$X_tX_{t-1} = \frac{2}{5}X_{t-1}^2 + \frac{1}{4}X_{t-2}X_{t-1} - \frac{1}{10}X_{t-3}X_{t-1} + Z_tX_{t-1}$

$X_tX_{t-2} = \frac{2}{5}X_{t-1}X_{t-2} + \frac{1}{4}X_{t-2}^2 - \frac{1}{10}X_{t-3}X_{t-2} + Z_tX_{t-2}$

$X_tX_{t-3} = \frac{2}{5}X_{t-1}X_{t-3} + \frac{1}{4}X_{t-2}X_{t-3} - \frac{1}{10}X_{t-3}^2 + Z_tX_{t-3}$

So we obtain,
(Note: $Cov(X_t,Z_t) = 0$)

$\gamma(1) = \frac{2}{5}\gamma(0) + \frac{1}{4}\gamma(1)-\frac{1}{10}\gamma(2)$

$\gamma(2) = \frac{2}{5}\gamma(1) + \frac{1}{4}\gamma(0)-\frac{1}{10}\gamma(1)$

$\gamma(3) = \frac{2}{5}\gamma(2) + \frac{1}{4}\gamma(1)-\frac{1}{10}\gamma(0)$

Which then gives,

$\rho(1) = \frac{2}{5} + \frac{1}{4}\rho(1)-\frac{1}{10}\rho(2)$

$\rho(2) = \frac{2}{5}\rho(1) + \frac{1}{4}-\frac{1}{10}\rho(1)$

$\rho(3) = \frac{2}{5}\rho(2) + \frac{1}{4}\rho(1)-\frac{1}{10}$

We get three equations and three unknowns,

$0.75\rho(1) + 0.1\rho(2) = 0.4$

$0.3\rho(1) - \rho(2) = -0.25$ 

$\rho(3) = 0.4\rho(2) + 0.25\rho(1)-0.1$

By rearranging (2), and substituting it in (1) we get,

$\rho(2) = 0.3\rho(1) + 0.25$  

$0.75\rho(1) + 0.1(0.3\rho(1) + 0.25) = 0.4$

$0.78\rho(1) = 0.375$

$\rho(1) = \frac{375}{780}$

which we can now use to find $\rho(2)$,

$\rho(2) = 0.3\rho(1) + 0.25$  

$\rho(2) = 0.3\big(\frac{375}{780}\big) + 0.25 = \frac{41}{104}$

Using both $\rho(1)$ and $\rho(2)$, we can solve for $\rho(3)$,

$\rho(3) = 0.4\rho(2) + 0.25\rho(1)-0.1$

$\rho(3) = 0.4\big(\frac{41}{104}\big) + 0.25\big(\frac{375}{780}\big)-0.1=\frac{37}{208}$

So, we have,

$\rho(1) = \frac{375}{780} \approx 0.481$

$\rho(2) = \frac{41}{104} \approx 0.394$

$\rho(3) = \frac{37}{208} \approx 0.178$

We know that $d_1, d_2, d_3$ are the roots of the polynomial,

$D^3 - 0.4D^2 - 0.25D + 0.1 = 0$

```{r cubic, include=TRUE}
polyroot(z = c(0.1, -0.25, -0.4, 1))
```

$d_1 = -0.5$, $d_2 = 0.4$, and $d_3 = 0.5$.

Thus we have,

$\rho(1) = A_1d_1 + A_2d_2 + A_3d_3$

$\rho(2) = A_1d_1^2 + A_2d_2^2 + A_3d_3^2$

$1 = A_1 + A_2 + A_3$

which becomes another system of 3 equations with 3 unknowns,

$-0.5A_1 + 0.4A_2 + 0.5A_3 = \frac{375}{780}$

$0.25A_1 + 0.16A_2 + 0.25A_3 = \frac{41}{104}$

$A_1 + A_2 + A_3 = 1$

We can solve the following using matrix functions,

```{r solving system of equations using matrix functions, include=TRUE}

a = matrix(c(-0.5, 0.4, 0.5, 0.25, 0.16, 0.25, 1, 1, 1), ncol = 3)
b = c(375/780, 41/104, 1)
x <- solve(t(a),b)
```

So, $A_1 = 0.179$, $A_2 = -1.603$ and $A_3 = 2.423$

Which finally gives us,

$\rho(h) = (0.179 \times (-0.5)^{|h|}) + (-1.603 \times (0.4)^{|h|}) + (2.423 \times (0.5)^{|h|})$


(c)

```{r sample and theoretical acf plots, include = TRUE}
set.seed(123)

z_t <- arima.sim(n = 2000, model = list(ar = c(0.4, 0.25, -0.1)), sd = sqrt(1.96))
a <- ARMAacf(ar = c(0.4, 0.25, -0.1))
acf(z_t, lag.max = 10)
points(c(0:3), a)
```

\pagebreak

## Question 2

(a) 

$X_t = \frac{7}{10}X_{t-1} + Z_t - \frac{1}{10}Z_{t-1}$

$X_t - \frac{7}{10}X_{t-1} = Z_t - \frac{1}{10}Z_{t-1}$

Therefore,

$\phi(B)X_t = (1 - 0.7B)X_t$

$\phi(B) = 1-0.7B$

For $\phi(B) = 0$, $B = \frac{1}{0.7} \approx 1.429 > 1$, so the process is stationary.

Similarly,

$\theta(B)Z_t = (1 - 0.1B)Z_t$

$\theta(B) = 1-0.1B$

For $\theta(B) = 0$, $B = \frac{1}{0.1} = 10 > 1$, so the process is invertible.

As all roots of $\phi(B) = 0$ and $\theta(B) = 0$ lie outside the unit circle of the complex plane, the process is stationary and invertible.

(b)

$X_t = \psi(B)Z_t$

$\psi(B) = \frac{\theta(B)}{\phi(B)} = \frac{1-0.1B}{1-0.7B} = (1-0.1B)(1+0.7B+(0.7B)^2...) = 1 + \sum_{j=1}^{\infty} (0.7^j - 0.1 \times 0.7^{j-1})B^j$

Hence,

$X_t = (1 + \sum_{j=1}^{\infty} (0.7^j - 0.1 \times 0.7^{j-1})B^j)Z_t = Z_t + \sum_{j=1}^{\infty} (0.7^j - 0.1 \times 0.7^{j-1})Z_{t-j}$

(c)

$\pi(B)X_t = Z_t$

$\pi(B) = \frac{\phi(B)}{\theta(B)} = \frac{1-0.7B}{1-0.1B} = (1-0.7B)(1+0.1B+(0.1B)^2+...) = 1 + \sum_{i=1}^{\infty} (0.1^i - 0.7 \times 0.1^{i-1})B^i$

Hence,

$Z_t = (1 + \sum_{i=1}^{\infty} (0.1^i - 0.7 \times 0.1^{i-1})B^i)X_t = X_t + \sum_{i=1}^{\infty} (0.1^i - 0.7 \times 0.1^{i-1})X_{t-i}$

(d)

From (b), we have the following result,
$X_t = Z_t + \sum_{j=1}^{\infty} (0.7^j - 0.1 \times 0.7^{j-1})Z_{t-j}$

which we can manipulate to give,

$X_t = Z_t + \sum_{j=1}^{\infty} (0.7^{j-1}(0.7 - 0.1))Z_{t-j}$

$X_t = Z_t + \sum_{j=1}^{\infty} (0.7^{j-1}(0.6))Z_{t-j}$

$X_t = Z_t + \sum_{j=1}^{\infty} 0.7^j \times \frac{6}{7} \times Z_{t-j}$

$acf(h) = \frac{\gamma_h}{\gamma_0} = \frac{Cov(X_t,X_{t+h})}{Var(X_t)} = \frac{(0.7^h)6^2\big(\frac{6}{7} + \frac{36}{49}\sum_{j=1}^{\infty}0.7^{2j}\big)}{6^2\big(1+\frac{36}{49}\sum_{j=1}^{\infty}0.7^{2j}\big)} = \frac{0.7^h\big(\frac{6}{7} + \big(\frac{36}{49} \times \frac{0.49}{1-0.49}\big)\big)}{1+\big(\frac{36}{49} \times \frac{0.49}{1-0.49}\big)} = \Big(\frac{186}{203}\Big)0.7^h \approx (0.916)0.7^h$

for $h>0$, (we know that when $h=0$, $acf(0)=1$)

\[ acf(h) = \begin{cases} 
      1 & h=0 \\
      \Big(\frac{186}{203}\Big)0.7^h & h>0 
   \end{cases}
\]

## Question 3

SARIMA $(2,1,0) \times (0,1,2)_{12}$

$(1-\phi_1B-\phi_2B^2)(X_t-X_{t-12}-X_{t-1}+X_{t-13}) = (1+\theta_1B^{12}+\theta_2B^{24})Z_t$

$X_t-X_{t-12}-X_{t-1}+X_{t-13}-\phi_1X_{t-1}+\phi_1X_{t-13}+\phi_1X_{t-2}-\phi_1X_{t-14}-\phi_2X_{t-2}+\phi_2X_{t-14}+\phi_2X_{t-3}-\phi_2X_{t-15}=Z_t+\theta_1Z_{t-12}+\theta_2Z_{t-24}$

$X_t-(1+\phi_1)X_{t-1}+(\phi_1-\phi_2)X_{t-2}+\phi_2X_{t-3}-X_{t-12}+(1+\phi_1)X_{t-13}-(\phi_1-\phi_2)X_{t-14}-\phi_2X_{t-15}=Z_t+\theta_1Z_{t-12}+\theta_2Z_{t-24}$

which gives us ARMA$(15,24)$.